PREVIC: An adaptive parent report measure of expressive vocabulary in children between 3 and 8 years of age

### 2025-02-20 paper/manuscript.Rmd
update paper link

affiliation:
  - id            : "1"
    institution   : "Institute for Psychology, Leuphana University Lüneburg, Germany"
  - id            : "2"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "3"
    institution   : "Institute of Psychology, Friedrich-Schiller-University Jena, Germany"
  - id            : "4"
    institution   : "Leibniz-Zentrum Allgemeine Sprachwissenschaft, Berlin, Germany"

@@

Child language measures can be broadly categorized into two types: direct and parent report measures. Direct measures of productive and receptive language are generally used with children of three years and older. Direct expressive language assessments involve prompting children to generate words or sentences in response to a stimulus, such as a picture or an object. Direct receptive language assessments reverse the logic and require children to match a verbal prompt with a picture or an object. Various direct measures tailored to different languages and age groups have been developed, including measures for English and German [@dunn1965peabody; @dunn1997british; @golinkoff2017user; @kiese2005awst; @kauschke2002patholinguistische; @lenhard2015peabody; @gluck2011wortschatz; @bohn2023orev; @armon2015assessing]. Additionally, standardized cognitive ability tests frequently incorporate direct language measures [e.g., @bayley2006bayley; @wechsler1949wechsler; @gershon2013iv].

@@

```{r figr, include = T, fig.align = "center", fig.cap = "(ref:figlab1)", out.width="50%"}
knitr::include_graphics("../graphs/figr.png")
```

@@

Our goal was to create an item pool with items of different word classes and varying semantical difficulty. We used Age-of-Acquisition (AoA) ratings as a rough indicator of anticipated item difficulty. Previous work has shown strong associations between AoA ratings and how likely children are to know a word [@bohn2023orev; @bohn2021young]. We started the process by compiling a list with AoA ratings for 3,921 German words from various sources [@luniewska2019age; @schroder2012german; @birchenough2017rated]. We excluded words with AoA ratings above ten. The remaining words were ordered by rated AoA and then split into ten lists with 344 words each. A research assistant with a background in linguistics went through the lists and selected words that a) were indicative of language abilities more broadly (avoiding very specialized terms) and b) that were different from one another in that they were semantically unrelated (to avoid words that are learned in the same context). For each list, we aimed for roughly 35 words from the three word classes; 17 nouns, nine verbs and nine adjectives. The so-generated item pool had `r data%>%distinct(word)%>%summarise(n())` words, of which `r words$noun` were nouns, `r words$verb` were verbs and `r words$adjective` were adjectives. Figure \@ref(fig:fig2)A shows how the items were distributed across AoA ratings and word types.

@@

In- and Outfit quantify the deviation of a person's response to an item from what the model predicts based on the difficulty of an item and the person's ability parameter. That is, both indices are a direct measure of how well the model was able to predict the responses to an item.   We fit a Rasch Model to the data and computed In- and Outfit values based on draws from the model’s expected posterior predictive distribution (using the function `add_epred_draws`) for each item and person combination. For each draw, we computed the residual between predicted and observed responses. The mean of the squared residuals is the Outfit; to obtain the Infit, the mean squared residuals are weighted by item information. The result was a distribution of values for each item and index. For each item, we then computed the mode for each index. The closer to 1 the index is, the better the fit. Figure \@ref(fig:fig3) visualizes the results. We used the cut-off values suggested in the literature [@bond2013applying; @debelak2022introduction] and excluded items with In- or Outfit values below 0.7 and above 1.3. Like all heuristics, these cut-offs are to some extent arbitrary. Yet, in the present context they served the purpose of removing a large number of potentially unsuitable items. This procedure led us to exclude `r fit_select$exclude` of the `r n_distinct(data$word)` items in the pool, leaving `r fit_select$include` for the automated item selection.

@@

First we defined an objective function that reflected the selection criteria which would later be used in the automated selection process. Items should vary in their difficulty but still cover all sections of the latent ability; we quantified this requirement as the standard deviation of the distance (in difficulty estimates) between adjacent items. The distance between adjacent items was computed by first sorting all items in the subset by difficulty and then subtracting the difficulty of adjacent items. Lower values indicate smaller distances and thus an overall more equal spacing. Items should also fit the Rasch model; we quantified this requirement in three ways. First and second, we used the In- and Outfit values for each item computed in the previous step. Third, we computed modification indices for each item. For this, we re-fitted the Rasch model using the package `lavaan` [@rosseel2012lavaan] and used the function `modindices` to obtain modification indices. Broadly speaking, modification indices quantify the improvement in model fit (in terms of the chi-square test statistic) when an item would be dropped [@rosseel2012lavaan].

@@

Following @bohn2023orev, we employed simulated annealing [@kirkpatrick1983optimization] as a method to identify the most optimal items for any given subset size. The process involves systematically exploring the vast space of possible subsets, commencing from a randomly selected initial subset. Subsequently, small random changes are proposed by exchanging some items within the subset under consideration with others located outside it. If a proposed change leads to an improvement in the objective function's value, the proposal is accepted, and the enhanced subset becomes the starting point for subsequent proposals.

@@

We assessed convergent validity in two studies. First, we compared PREVIC scores to a direct assessment of children's receptive vocabulary using the oREV [@bohn2023orev]. The oREV asks children to select a picture (out of four) upon hearing a word. It has 22 items which fit the Rasch model. Because the oREV is also available as a web application, we sent out emails to all caregivers who provided complete data in the data collection that led to the construction of the PREVIC (N = `r n_distinct(data$subjID)`) and asked them to have their child complete the oREV. We obtained oREV data from `r link_dem$n` children (`r link_dem$fem` female, $m_{age}$ = `r link_dem$age_mean`, range = `r link_dem$age_min` - `r link_dem$age_max`) which corresponds to a response rate of \~ `r round(692/1190*100)`%. We found a substantial correlation between caregiver's answers to questions about their children's expressive vocabulary in the PREVIC and a direct assessment of children's receptive vocabulary in the oREV (*r* = `r round(val_cor$estimate,2)`; 95% CI = `r round(val_cor$conf.int[1],2)` -- `r round(val_cor$conf.int[2],2)`; Figure \@ref(fig:fig4)B).

@@

When assessing convergent validity, we found a somewhat lower correlation between PREVIC scores and a direct measure of children’s expressive vocabulary (AWST-R) compared to receptive vocabulary (oREV). Potential reasons for this pattern could be as follows: the AWST-R has not been revised in nearly 20 years, and some of its images may now seem outdated (e.g., "rauchen" (Eng. "smoking") or "telefonieren" (Eng. “to phone”)), making them less suited to assess expressive everyday language. Relatedly, mean performance in the validation study was near the upper end of the AWST-R score range (though not at ceiling), which made it challenging to discriminate between individuals due to the scarcity of very difficult items. Another possible reason is the use of a relatively large standard error (0.4) as the criterion for terminating the PREVIC. On average, parents responded to only around 34 items (out of 89), often completing them in under five minutes. In sum, these factors may have led to less precise measurement at both ends of the scale, potentially contributing to the relatively lower correlation. We therefore recommend to use a smaller standard error of e.g., 0.3 for the adaptive version. Note, however, that a standard error of 0.2 usually leads to a presentation of all items in the pool.

>>>>

affiliation:
  - id            : "1"
    institution   : "Institute of Psychology in Education, Leuphana University Lüneburg, Germany"
  - id            : "2"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "3"
    institution   : "Institute of Psychology, Friedrich-Schiller-University Jena, Germany"
  - id            : "4"
    institution   : "Leibniz-Zentrum Allgemeine Sprachwissenschaft, Berlin, Germany"

@@

Child language measures can be broadly categorized into two types: direct and parent report measures. Direct measures of productive and receptive language are generally used with children of three years and older. Direct expressive language assessments involve prompting children to generate words or sentences in response to a stimulus, such as a picture or an object. Direct receptive language assessments reverse the logic and require children to match a verbal prompt with a picture or an object. Various direct measures tailored to different languages and age groups have been developed, including measures for English and German [@dunn1965peabody; @dunn1997british; @golinkoff2017user; @kiese2005awst; @kauschke2002patholinguistische; @lenhard2015peabody; @gluck2011wortschatz; @bohn2024orev; @armon2015assessing]. Additionally, standardized cognitive ability tests frequently incorporate direct language measures [e.g., @bayley2006bayley; @wechsler1949wechsler; @gershon2013iv].

@@

```{r figr, include = T, fig.align = "center", fig.cap = "(ref:figlabr)", out.width="50%"}
knitr::include_graphics("../graphs/figr.png")
```

@@

Our goal was to create an item pool with items of different word classes and varying semantical difficulty. We used Age-of-Acquisition (AoA) ratings as a rough indicator of anticipated item difficulty. Previous work has shown strong associations between AoA ratings and how likely children are to know a word [@bohn2024orev; @bohn2021young]. We started the process by compiling a list with AoA ratings for 3,921 German words from various sources [@luniewska2019age; @schroder2012german; @birchenough2017rated]. We excluded words with AoA ratings above ten. The remaining words were ordered by rated AoA and then split into ten lists with 344 words each. A research assistant with a background in linguistics went through the lists and selected words that a) were indicative of language abilities more broadly (avoiding very specialized terms) and b) that were different from one another in that they were not highly semantically related (to avoid words that are learned in the same context). For each list, we aimed for roughly 35 words from the three word classes; 17 nouns, nine verbs and nine adjectives. The so-generated item pool had `r data%>%distinct(word)%>%summarise(n())` words, of which `r words$noun` were nouns, `r words$verb` were verbs and `r words$adjective` were adjectives. Figure \@ref(fig:fig2)A shows how the items were distributed across AoA ratings and word types.

@@

In- and Outfit quantify the deviation of a person's response to an item from what the model predicts, based on the difficulty of an item and the person's ability parameter. That is, both indices are a direct measure of how well the model was able to predict the responses to an item. We fit a Rasch Model to the data and computed In- and Outfit values based on draws from the model’s expected posterior predictive distribution (using the function `add_epred_draws`) for each item and person combination. For each draw, we computed the residual between predicted and observed responses. The mean of the squared residuals is the Outfit; to obtain the Infit, the mean squared residuals are weighted by item information (i.e., the probability of responding yes multiplied by the probability of answering no). The result was a distribution of values for each item and index. For each item, we then computed the mode for each index. The closer to 1 the index is, the better the fit. Figure \@ref(fig:fig3) visualizes the results. We used the cut-off values suggested in the literature [@bond2013applying; @debelak2022introduction] and excluded items with In- or Outfit values below 0.7 and above 1.3. Like all heuristics, these cut-offs are to some extent arbitrary. Yet, in the present context they served the purpose of removing a large number of potentially unsuitable items. This procedure led us to exclude `r fit_select$exclude` of the `r n_distinct(data$word)` items in the pool, leaving `r fit_select$include` for the automated item selection.

@@

First we defined an objective function that reflected the selection criteria which would later be used in the automated selection process. Items should vary in their difficulty but still cover all sections of the latent ability; we quantified this requirement as the standard deviation of the distance (in difficulty estimates) between adjacent items. The distance between adjacent items was computed by first sorting all items in the subset by difficulty and then subtracting the difficulty of adjacent items. A lower standard deviation of these distances indicates little variation between them and thus an overall more equal spacing. Items should also fit the Rasch model; we quantified this requirement in three ways. First and second, we used the In- and Outfit values for each item computed in the previous step. Third, we computed modification indices for each item. For this, we re-fitted the Rasch model using the package `lavaan` [@rosseel2012lavaan] and used the function `modindices` to obtain modification indices. Broadly speaking, modification indices quantify the improvement in model fit (in terms of the chi-square test statistic) when an item would be dropped [@rosseel2012lavaan].

@@

Following @bohn2024orev, we employed simulated annealing [@kirkpatrick1983optimization] as a method to identify the most optimal items for any given subset size. The process involves systematically exploring the vast space of possible subsets, commencing from a randomly selected initial subset. Subsequently, small random changes are proposed by exchanging some items within the subset under consideration with others located outside it. If a proposed change leads to an improvement in the objective function's value, the proposal is accepted, and the enhanced subset becomes the starting point for subsequent proposals.

@@

We assessed convergent validity in two studies. First, we compared PREVIC scores to a direct assessment of children's receptive vocabulary using the oREV [@bohn2024orev]. The oREV asks children to select a picture (out of four) upon hearing a word. It has 22 items which fit the Rasch model. Because the oREV is also available as a web application, we sent out emails to all caregivers who provided complete data in the data collection that led to the construction of the PREVIC (N = `r n_distinct(data$subjID)`) and asked them to have their child complete the oREV. We obtained oREV data from `r link_dem$n` children (`r link_dem$fem` female, $m_{age}$ = `r link_dem$age_mean`, range = `r link_dem$age_min` - `r link_dem$age_max`) which corresponds to a response rate of \~ `r round(692/1190*100)`%. We found a substantial correlation between caregiver's answers to questions about their children's expressive vocabulary in the PREVIC and a direct assessment of children's receptive vocabulary in the oREV (*r* = `r round(val_cor$estimate,2)`; 95% CI = `r round(val_cor$conf.int[1],2)` -- `r round(val_cor$conf.int[2],2)`; Figure \@ref(fig:fig4)B).

@@

When assessing convergent validity, we found a somewhat lower correlation between PREVIC scores and a direct measure of children’s expressive vocabulary (AWST-R) compared to receptive vocabulary (oREV). Potential reasons for this pattern could be as follows: the AWST-R has not been revised in nearly 20 years, and some of its images may now seem outdated (e.g., "rauchen" (Eng. "smoking") or "telefonieren" (Eng. “to phone”)), making them less suited to assess expressive everyday language. Relatedly, mean performance in the validation study was near the upper end of the AWST-R score range (though not at ceiling), which made it challenging to discriminate between individuals due to the scarcity of very difficult items. Another possible reason is the use of a relatively large standard error (0.4) as the criterion for terminating the PREVIC. On average, parents responded to only around 34 items (out of 89), often completing them in under five minutes. A smaller standard error of e.g. 0.3 would have resulted in a longer test (more items) and a more precise estimate of the person ability. In sum, these factors may have led to less precise measurement at both ends of the scale, potentially contributing to the relatively lower correlation. We therefore recommend to use a smaller standard error of e.g., 0.3 for the adaptive version. Note, however, that a standard error of 0.2 usually leads to a presentation of all items in the pool.
